Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-01 documents/ms.tex
tidying up a bit, adding more words to the conclusion

We calculated an ACF for each light curve following the method of
\citet{Mcquillan2013}.
An ACF is calculated for each light curve and smoothed by convolving with a
Gaussian with $\sigma=9$ days.
A rotation period is estimated as the lag-time of the highest peak in the ACF,
less than 100 days.
This is not always the first peak: the second can be larger than the first if
there are two active regions at or near opposite longitudes on the surface of
the star---producing a light curve with two dips per rotation period.
We restrict our search to periods less than 100 days because this is the
maximum period that is realistically recoverable in \kepler\ light curves ---
instrumental noise significantly distorts signals on longer timescales.
An example ACF of the light curve in figure~\ref{fig:demo_lc} is shown
in figure \ref{fig:demo_acf}.

@@

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2n$ and $\frac{1}{2}n$ harmonic lines are shown as dashed lines.
The Median Absolute Deviation (MAD) of the ACF-recovered periods is \acfRMS\
in log-space (1.8 days).
The agreement between the injected and recovered rotation periods is good in
general, however twice the true rotation period is often found.
In addition, some rotation periods are drastically over- and underestimated.
Of the points clustered close to the $x=y$ line, more fall slightly below it
than above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves are similar to the QP kernel function in equation
\ref{eq:QP}: periodic functions added to decaying exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and that used largely in the literature.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_acf.pdf}
\caption[ACF results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using the ACF technique.
    Several light curves have over-estimated rotation periods and some
    are drastically underestimated.
    In some cases the recovered period is twice the true period used to
    simulate the light curve.}
\label{fig:compare_acf}
\end{center}
\end{figure*}

@@

For each simulated light curve, a LS periodogram
\footnote{LS periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
was computed over a grid of 10,000 periods, evenly spaced in frequency,
between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
The MAD for rotation periods measured using the LS periodogram technique is
\pgramRMS\ in log-space (2.7 days).
% We believe that the unsuccessful recoveries occurring preferentially for
% shorter rotation periods are produced by additional, longer timescale
% variations in some of the light curves.
% Specifically, by fluctuations produced by changes in the total spot coverage
% on the stellar surface.
% These fluctuations can occasionally be larger in magnitude than those produced
% by the rotational signal itself and are present in the majority of outlying
% light curves that occupy the top left of figure
% \ref{fig:pgram_compare_noise-free}.

@@

Figures \ref{fig:compare_mcmc_noprior} and \ref{fig:compare_mcmc_acfprior}
summarize the results of this MCMC fitting procedure compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
Both the informative and uninformative prior versions of the GP method perform
better than the ACF and periodogram methods, with MADs of 0.38 and 0.48,
respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/25.png}
\caption{Marginal posterior PDFs of the QP GP model parameters. $\sigma$ is an
additional white noise term added to the diagonal elements of the covariance
matrix. It is the fraction by which the observational uncertainties have been
underestimated. If the uncertainties reported on the data are too small, this
parameter will be non-zero.
    The green line in the period panel shows the injected period.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

% At this point it is important to highlight the following caveat.
% Our method has been tuned to reproduce the rotation periods of the
% \citet{Aigrain2015} simulated light curves.
% These light curves were simulated using a very good but still imperfect model.
% The input parameters and their distributions were chosen to be as realistic as
% possible given current observations.
% Unfortunately however, there are limited data on some of these parameters for
% stars other than the Sun: for example the distrubution of spot lifetimes or
% activity cycle length.
% Ensuring that we can recover the rotation periods of these simulated light
% curves does not necessarily ensure that we can recover the rotation periods of
% real \kepler\ light curves with equal accuracy and precision.
% Even using priors on the QP kernel parameters that have been derived from the
% \citet{Aigrain2015} sample may be dangerous.
% This caveat should be taken into consideration when this method is applied to
% real data.

@@

The uncertainties produced by the GP method are more representative than those
produced by the ACF and periodogram methods because the posterior PDF of the
period is explored using MCMC.
Whilst they are an improvement, unfortunately these uncertainties are still
underestimated in the majority cases.
Of the rotation periods recovered from the simulated light curves, only 25\%
of the measured periods lay within 1$\sigma$ of the true period.
50\% lay within 2$\sigma$ and 66\% within 3$\sigma$.
The largest outlier was 114$\sigma$ away from the true value.
In the majority of these cases the uncertainties are underestimated due to the
multi-modal nature of the period posterior PDF.
Because the posteriors are multi-modal they are difficult to sample and,
without running the MCMC for infinitely long, the samples will not necessarily
always cover the full range of periods with significant power.
The underestimated uncertainties can also be attributed to the model choice.
Although a GP model is more appropriate than a sinusoid, it is still an
imperfect, effective model.
A perfect physical model of the star would produce better results with more
representative uncertainties.

The QP kernel function represents a simplistic effective model of a stellar
light curve.
It adequately describes the data, captures that all-important periodic quality
and is relatively simple, with only a few hyper-parameters.
It also satisfies the requirement to produce positive semi-definite covariance
matrices.
Whilst the QP kernel function evidently captures the periodic qualities of
light curves adequately, it is still a somewhat arbitrary choice.
Another valid choice would be a squared cosine function multiplied by a
squared exponential,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2}\right)
\cos\left(\frac{2\pi}{P}\right)
\end{equation}
\label{eq:cos_kernel}
This function produces a positive semi-definite matrix and has the $P$
parameter of interest.
It may in fact be even {\it more} suited to modelling stellar
light curves as it describes a Gaussian in frequency space.
It is easy to imagine a differentially rotating star with a period that is a
Gaussian in frequency space: the mean frequency would be the frequency at the
most active latitude, at or near which spots spend the majority of their time
and the tails would be occupied by spots that drift near the equator or poles.
The main difference between this cosine and the QP function is that the cosine
function allows negative covariances and the QP function does not.
Is is realistic to allow negative covariances?
In practice, the ACFs of \Kepler\ light curves often go negative.
However, many stars have two active regions on opposite hemispheres that
produce two brightness dips per rotation.
If the covariance is forced to be negative for two data points that are
separated by half a rotation period, those light curves with two peaks per
rotation period may not be well modelled.
In other words, we do not want to force anti-correlation of points that are
$\frac{1}{2}$ a period apart.
It would be very worthwhile to test this assumption and this alternative
kernel function in future.
If CPU time were not limited there may be some benefit to performing formal
model comparison with different kernel functions.
However, the evidence integral is an ambitious calculation for models with
likelihood functions that take milliseconds to compute, let alone those
involving GPs with light curves containing thousands of data points which can
take minutes.

Clearly, stellar rotation periods are well represented by the $P$ parameter in
our QP kernel function, as evidenced by the its impressive ability to recover
the true rotation periods from the simulated light curves in this work.
However, it may not be the {\it best} function to use.
It is, after all, an approximation to the shape of the covariance matrix of a
real stellar light curve.
There may be an alternative function which is better suited to capturing
stellar variability and is able to recover periods even more precisely than
the QP kernel.
There may also be an alternative function which is more physically motivated,
that captures not just the rotation period but also (for example) the spot
lifetime or differential surface rotation.
This is beyond the scope of this work but we hope that these questions will be
answered in the near future.

@@

\item{To develop a detection criterion to assess whether a rotation period was
measured.
Another general problem in rotation period inference is deciding whether a
real rotation period was measured at all.
Detection thresholds are usually set using the amplitudes of peaks in LS
periodograms or ACFs in order to eliminate contaminants.
Using a GP model it may be possible to do this via model selection: \ie\ are
the data better described by a periodic model, or a non-periodic model?
There is more than one way to implement such a model comparison, one way could
be to use a sum of two kernel functions: one periodic and one non-periodic,
each with its own amplitude parameter.}

@@

\section{Conclusions}

We used the \naigrain\ simulated \kepler-like light curves for solid-body
rotation from \citet{Aigrain2015} and attempted to recover the rotation
periods used to generate them.
Three methods were compared: a Lomb-Scargle periodogram method, an
autocorrelation function method and a new Gaussian process method.
We demonstrate that the GP method produced the most precise and accurate
rotation periods of the three techniques.
As expected, it provides a large improvement over the periodogram method and a
moderate one over the ACF method.
The periodogram method performs poorly because a sinusoid is not a perfect
model choice for stellar light curves --- they do not always display
sinusoidal behaviour.
On the other hand, the ACF method does not rely on any model choice.

In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2014}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data.

The GP method provides uncertainties because it produces posterior PDF samples
but they are still imperfect estimates of the intrinsic uncertainties.
This is because the posterior PDF of the rotation period parameter is, in many
cases, multi-modal.
If it were possible to run an MCMC sampler for an infinite number of steps,
the entire posterior PDF would eventually be sampled, producing an accurate
uncertainty measurement.
Practically speaking however, these imperfect estimates may be the best
currently achievable.

A GP is a semi-parametric model that is well suited to signals that are
non-sinusoidal and quasi-periodic.
Not only this, unlike the other two methods, the GP method provides posterior
PDF samples of rotation periods, allowing characterisation of uncertainties.
The posterior PDFs are often non-Gaussian, being bi- or multimodal.
Rather, it is better to incorporate posterior PDF samples into analyses which
require uncertainty propagation.
Because rotation periods are inferred probabilistically with the GP method,
this allows them to be easily incorporated into follow-on studies; inferring
stellar ages or characterising star-planet interactions for example.

>>>>

We measured an ACF-based period for each light curve following the method of
\citet{Mcquillan2013}.
An ACF is calculated for each light curve and smoothed by convolving with a
Gaussian with $\sigma=9$ days.
A rotation period is estimated as the lag-time of the highest peak in the ACF,
less than 100 days.
This is not always the first peak: the second can be larger than the first if
there are two active regions at or near opposite longitudes on the surface of
the star---producing a light curve with two dips per rotation period.
We restrict our search to periods less than 100 days because this is the
maximum period that is realistically recoverable in \kepler\ light curves ---
instrumental noise significantly distorts signals on longer timescales.
An example ACF of the light curve in figure~\ref{fig:demo_lc} is shown
in figure \ref{fig:demo_acf}.

@@

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2n$ and $\frac{1}{2}n$ harmonic lines are shown as dashed lines.
The Median Absolute Deviation (MAD) of the ACF-recovered periods is \acfRMS\
in log-space (1.8 days).
The agreement between the injected and recovered rotation periods is good in
general, however some rotation periods are drastically over- and
underestimated.
Of the points clustered close to the $x=y$ line, more fall slightly below it
than above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and that used largely in the literature.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_acf.pdf}
\caption[ACF results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using the ACF technique.
    Several light curves have over-estimated rotation periods and some
    are drastically underestimated.}
\label{fig:compare_acf}
\end{center}
\end{figure*}

@@

For each simulated light curve, a LS periodogram
\footnote{LS periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
was computed over a grid of 10,000 periods, evenly spaced in frequency,
between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
The MAD for rotation periods measured using the LS periodogram technique is
\pgramRMS\ in log-space (2.7 days).
There is a large number of drastically overestimated periods recovered using
the periodogram.
Excess power at long periods can be generated by long-term trends that are not
related to rotation.
This problem becomes exacerbated when instrumental noise is also present in
light curves.

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarize the results of this MCMC fitting procedure compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
Both the informative and uninformative prior versions of the GP method perform
better than the ACF and periodogram methods, with respective MADs of 0.38 and
0.48, compared to \acfRMS\ (ACF) and \pgramRMS\ (periodogram).
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/25.png}
\caption{Marginal posterior PDFs of the QP GP model parameters, fit to the
    light curve in figure \ref{fig:demo_lc}.
$\sigma$ is an
additional white noise term added to the diagonal elements of the covariance
matrix. It is the fraction by which the observational uncertainties have been
underestimated. If the uncertainties reported on the data are too small, this
parameter will be non-zero.
    The green line in the period panel shows the injected period.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

@@

The uncertainties produced by the GP method are more representative than those
produced by the ACF and periodogram methods because the posterior PDF of the
period is explored using MCMC.
Whilst they are an improvement, unfortunately these uncertainties are still
underestimated in the majority cases.
Of the rotation periods recovered from the simulated light curves, only 25\%
of the measured periods lie within 1 $\sigma$ of the true period.
50\% lie within 2 $\sigma$ and 66\% within 3 $\sigma$.
The largest outlier is 114 $\sigma$ away from the true value.
In the majority of these cases the uncertainties are underestimated due to the
multi-modal nature of the period posterior PDFs which makes them difficult to
sample.
The underestimated uncertainties can also be attributed to the model choice.
Although a GP model is more appropriate than a sinusoid, it is still an
imperfect, effective model.
A perfect physical model of the star would produce better results with more
representative uncertainties.

The QP kernel function represents a simplistic effective model of a stellar
light curve.
It adequately describes the data, captures that all-important periodic quality
and is relatively simple, with only a few hyper-parameters.
It also satisfies the requirement to produce positive semi-definite covariance
matrices.
Whilst the QP kernel function evidently captures the periodic qualities of
light curves adequately, it is still a somewhat arbitrary choice.
Another valid choice would be a squared cosine function multiplied by a
squared exponential,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2}\right)
\cos\left(\frac{2\pi}{P}\right)
\end{equation}
\label{eq:cos_kernel}
This function produces a positive semi-definite matrix and has the $P$
parameter of interest.
% It may in fact be even {\it more} suited to modelling stellar
% light curves as it describes a Gaussian in frequency space.
% It is easy to imagine a differentially rotating star with a period that is a
% Gaussian in frequency space: the mean frequency would be the frequency at the
% most active latitude, at or near which spots spend the majority of their time
% and the tails would be occupied by spots that drift near the equator or poles.
The main practical difference between this cosine and the QP function is that
the cosine function allows negative covariances and the QP function does not.
Is is realistic to allow negative covariances?
In practice, the ACFs of \Kepler\ light curves often go negative.
However, many stars have two active regions on opposite hemispheres that
produce two brightness dips per rotation.
If the covariance is forced to be negative for two data points that are
separated by half a rotation period, those light curves with two peaks per
rotation period may not be well modelled.
In other words, we do not want to force anti-correlation of points that are
$\frac{1}{2}$ a period apart.
It would be very worthwhile to test this assumption and this alternative
kernel function in future.
If CPU time were not limited there may be some benefit to performing formal
model comparison with different kernel functions.
However, the evidence integral is an ambitious calculation for models with
likelihood functions that take milliseconds to compute, let alone those
involving GPs with light curves containing thousands of data points which can
take minutes.

% Clearly, stellar rotation periods are well represented by the $P$ parameter in
% our QP kernel function, as evidenced by the its impressive ability to recover
% the true rotation periods from the simulated light curves in this work.
% However, it may not be the {\it best} function to use.
% It is, after all, an approximation to the shape of the covariance matrix of a
% real stellar light curve.
% There may be an alternative function which is better suited to capturing
% stellar variability and is able to recover periods even more precisely than
% the QP kernel.
% There may also be an alternative function which is more physically motivated,
% that captures not just the rotation period but also (for example) the spot
% lifetime or differential surface rotation.
% This is beyond the scope of this work but we hope that these questions will be
% answered in the near future.

@@

% \item{To develop a detection criterion to assess whether a rotation period was
% measured.
% Another general problem in rotation period inference is deciding whether a
% real rotation period was measured at all.
% Detection thresholds are usually set using the amplitudes of peaks in LS
% periodograms or ACFs in order to eliminate contaminants.
% Using a GP model it may be possible to do this via model selection: \ie\ are
% the data better described by a periodic model, or a non-periodic model?
% There is more than one way to implement such a model comparison, one way could
% be to use a sum of two kernel functions: one periodic and one non-periodic,
% each with its own amplitude parameter.}

@@

\section{Conclusion}

We used the \naigrain\ simulated \kepler-like light curves for solid-body
rotation from \citet{Aigrain2015} and attempted to recover the rotation
periods used to generate them.
Three methods were compared: a Lomb-Scargle periodogram method, an
autocorrelation function method and a new Gaussian process method.
We demonstrate that the GP method produced the most precise and accurate
rotation periods of the three techniques.
As expected, it provides a large improvement over the periodogram method and a
moderate one over the ACF method.
We find that the ACF method often slightly under-predicts the rotation period
due to a subtlety in the standard peak detection algorithm.
% The periodogram method performs poorly because a sinusoid is not a perfect
% model choice for stellar light curves --- they do not always display
% sinusoidal behaviour.
% On the other hand, the ACF method does not rely on any model choice and
In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2014}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data.

The GP method provides approximate uncertainties because it produces posterior
PDF samples but they are still imperfect estimates of the intrinsic
uncertainties.
This is because the posterior PDF of the rotation period parameter is, in many
cases, multi-modal.
If it were possible to run an MCMC sampler for an infinite number of steps,
the entire posterior PDF would eventually be sampled, producing an accurate
uncertainty measurement.
We find that only a quarter of the reported uncertainties are representative.

The main aim of this work was to develop a rotation period inference method
that is probabilistic.
Probabilistic periods methods are necessary for hierarchical Bayesian
inference and for performing population analysis.
Using this method to generate a catalogue of probabilistic rotation periods
could, for example, reveal the period distribution of (and therefore
potentially the age distribution) of stars in the Milky Way.
There is still some work to be done on the development of this model such that
the uncertainty estimates are truly representative, but this new method
provides a improvement upon previous methods in that regard.
